---
title: "A Crash Course in Everything"
subtitle: "Git, Quarto, and R"
author: Dr. Gilbert
format: 
  revealjs:
    smaller: true
date: today
date-format: long
theme: serif
incremental: true
---

```{r global-options, include=FALSE}
library(tidyverse)
library(tidymodels)
library(countdown)

theme_set(theme_bw(base_size = 20))
```

```{css}
code.sourceCode {
  font-size: 1.3em;
  /* or try font-size: xx-large; */
}

a:link{
  color: purple;
}

a:visited{
  color: purple;
}
```

## Objectives

These slides address the following items.

  + How do I create a GitHub repo?
  + How do I connect my GitHub repo to an R Project on my local machine?
  + What is a Quarto Document and how do I use it?
  + How do I install and load packages in R? In particular we'll work with the `{tidyverse}`.
  + How do I read data into R from both local and remote sources?
  + How do I interact with, and manipulate, data using the tools and principles of the `{tidyverse}`?
  
## Creating a GitHub Repo

+ Navigate to [GitHub](https://www.github.com) and log in
+ Click the green button labeled `New` 
+ Name your repo (something descriptive, like *MAT434* or *ClassificationCourse*)
+ Click the checkbox to initialize the repo with a README file
+ Click the green button to Create repository.

## Cloning Your Repo to a Local R Project

+ Inside your GitHub repo, locate the green *Code* dropdown
+ Click it and copy the HTTPS address
+ Open RStudio
+ Use `File -> New Project` to begin creating a new R project
+ Select *From Version Control*, then *Git*, and paste in the URL
+ Hit the *Browse* button to change where the project will live
+ Navigate to *Documents*, click to create a new folder: *GitHub*
+ Hit *Open*/*Okay* and then click *Create*

## Create a New Quarto Document

+ Use `File -> New File -> QUarto Document` to create a new Quarto notebook
+ Add a title -- something like *Crash Course in Git, Quarto, and R*
+ Add yourself as the author
+ Click *Create*
+ Use `File -> Save As...` to save the file -- give it a name like `CrashCourse.qmd` and leave it in the project directory
+ Click the blue arrow button to *render* the notebook to HTML

## Back to Git...

+ Click on the *Git* tab in the top-right pane of RStudio
+ You've got some yellow flagged files listed there -- these are files not currently in your *remote* repo
+ Click the *Pull* button (you're already up to date, but it's a good practice to get in)
+ Click the checkboxes next to all your new files
+ Click the *Commit* button, add a message in the message box, and click *Commit*
+ Hit the push button to send your files out to the remote repo

. . . 

Head back to your web browser and refresh your repo -- your files have arrived!

## The Quarto Notebook Environment

+ Quarto documents consist of three main components 

  + a YAML header
  + code chunks (grey background)
  + text/markdown cells (white background)

+ We'll generally copy/paste the same YAML header (I'll give you a slightly more complex template)
+ Code chunks must have executable R code in them
+ Text/markdown cells can have anything you like

. . . 

Here's [a Quarto cheatsheet](https://rstudio.github.io/cheatsheets/html/quarto.html), but feel free to work with your favorite LLM to get Quarto to do exactly what you want

## Working in R

For the rest of the slide deck, we'll transition to working in R but then come back to Git/GitHub at the end of the class meeting

. . . 

**Students coming from MAT300:** Feel free to grab the data and start investigating

+ MLB [batted balls](https://raw.githubusercontent.com/agmath/agmath.github.io/refs/heads/master/data/classification/battedballs.csv) and [park dimensions](https://raw.githubusercontent.com/agmath/agmath.github.io/refs/heads/master/data/classification/park_dimensions.csv), or
+ FAA [wildlife impacts with planes](https://raw.githubusercontent.com/agmath/agmath.github.io/refs/heads/master/data/classification/faa_birdstrikes.csv)

. . . 

**Question for New Students:** Do you want to partner up with an existing student to work and learn from them, or do you want to continue with the slides?

## The R Dialects

There are three main dialects in R

+ Base-R
+ `data.table` (speed)
+ Tidy-R / `tidyverse` (readability and consistency)

  + We use this one. When you search for help -- include "*tidyverse*" in your query.
  
. . .

> **Note:** *R dialects* just refers to how we choose to write R code and which functions we prioritize -- dialects can be (and often are) mixed.

## Installing and Loading Packages

+ `install.packages("PACKAGE_NAME")` to install a package

  + You only do this once
  + Do it in the *console*

+ `library(PACKAGE_NAME)` to load the package

  + Do this near the top of your Quarto Document, in a *setup* chunk
  
. . .

$\bigstar$ Install the `{tidyverse}` and load it in your Quarto Notebook

. . . 

$\bigstar$ While you are at it, install the `{skimr}`, `{tidymodels}`, `{patchwork}`, and `{kableExtra}` packages as well -- `{skimr}` and `{kableExtra}` are the only ones you might want to load for now

## Reading in Data

. . .

**ASIDE:** We store objects in R with the arrow operator (`<-`)

. . .

```{r}
#| echo: true
#| eval: false

x <- 2
```

. . .

**Reading Data:** To read data, we use a function of the form `read_*()`.

. . .

```{r}
#| echo: true
#| eval: false

data <- read_csv("PATH_TO_CSV_FILE")
```

+ Requires the `{tidyverse}` (or at least `{readr}`) to be loaded
+ Similar functions exist for reading other file formats
+ Some require other packages (`{readxl}` or `{haven}` are common)

. . .

$\bigstar$ Read the MLB [batted balls data](https://raw.githubusercontent.com/agmath/agmath.github.io/refs/heads/master/data/classification/battedballs.csv) and the [park dimensions data](https://raw.githubusercontent.com/agmath/agmath.github.io/refs/heads/master/data/classification/park_dimensions.csv) into your Quarto Notebook

. . .

I'll post the links in Slack

## First Interactions with Data

+ `head()` to view first six rows
+ `glimpse()` to view dimensions and data types
+ `skim()` from `{skimr}` for much more detail

. . . 

$\bigstar$ Try these functions on your `battedballs` data

```{r}
countdown(minutes = 2)
```

## The Pipe Operator 

+ Pipes (`%>%` or `|>`) make code more readable and allow chaining of functions together
+ Object to the left of the pipe becomes first argument to the function after the pipe
+ Read the pipe to mean "*and then*"

. . .

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1|2|"

penguins %>%
  head()
```

. . . 

$\bigstar$ Rewrite the functions you used to explore your data with *pipes*

## Manipulating and Transforming Data

+ `filter()` to return only desired records
+ `select()` to return only desired columns
+ `summarize()` to compute summaries on a table
+ `group_by()` to create groups in a table
+ `mutate()` to create new columns or change existing ones

. . .

$\bigstar$ How might we use these functions? Write down some questions that could be answered using the functions described above. Start with a couple very simple questions and then work up to questions whose answers might be more complex to find.

## Manipulating and Transforming Data

:::{.nonincremental}

+ `filter()` to return only desired records
+ `select()` to return only desired columns
+ `summarize()` to compute summaries on a table
+ `group_by()` to create groups in a table
+ `mutate()` to create new columns or change existing ones

:::

$\bigstar$ How might we use these functions? Write down some questions that could be answered using the functions described above. Start with a couple very simple questions and then work up to questions whose answers might be more complex to find.

$\bigstar$ We'll try answering some of those questions now!

## All Changes are Temporary

```{r}
#| echo: true
#| eval: false

penguins %>%
  filter(species == "Gentoo") %>%
  group_by(island) %>%
  summarize(
    avg_mass = mean(body_mass_g)
  )
```

+ Start with the `penguins` data frame, *and then*
+ filter to just the Gentoo species, *and then*
+ group by island, *and then*
+ calculate average penguin body mass for each group

**Note:** `penguins` data frame is not permanently altered here

## Until We Make Them Permanent

. . .

```{r}
#| echo: true
#| eval: false

penguins <- penguins %>%
  filter(species == "Gentoo") %>%
  group_by(island) %>%
  summarize(
    avg_mass = mean(body_mass_g)
  )
```

. . .

Now the change is permanent because we've stored the result

+ Notice the use of the arrow operator (`<-`)
+ Be careful overwriting existing objects -- think about whether you:
  
  + might need the old object again
  + would be better off creating a new object (variable)
  + even need to store the result at all
    
## Let's Practice

Use this time to continue playing with the MLB data sets

1. Use the blue *render* button to convert your markdown document into a beautiful HTML document and enjoy the fruits of your labor!

    + 🤔 Ponder an existence where you never need to open MS Word (or PowerPoint) again! 🤔
    
1. Write down and answer additional interesting questions that might use functionality discussed in this slide deck -- start simple and then build up to questions that might be more complex
1. Document your work by including text descriptions alongside the code chunks

. . .

Don't worry if your document looks quite plain for now, our next class meeting is devoted to using markdown syntax in Quarto effectively

## Next Time...

<center>

<font size="120pt"><br/>Quarto, Inline R, and semi-automated reporting</font>

</center>