---
title: "Data Viz and `ggplot()` Intro"
author: Dr. Gilbert
format: 
  revealjs:
    smaller: true
date: today
date-format: long
theme: serif
incremental: true
---

```{r global-options, include=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
library(countdown)

options(kable_styling_bootstrap_options = c("hover", "striped"))

theme_set(theme_bw(base_size = 20))
```

```{css}
code.sourceCode {
  font-size: 1.3em;
  /* or try font-size: xx-large; */
}

a:link{
  color: purple;
}

a:visited{
  color: purple;
}
```

## The Highlights

+ Sample Viz
+ Bad Viz
+ Suggested Viz Choices
+ Structure of a `ggplot()`
+ Suggested Viz Choices, Revisited
+ On Plot Labels
+ Organizing plots with `{patchwork}`

## A Good Viz Tells a Clear Story

```{r}
#| echo: false
#| message: false
#| warning: false

p1 <- mpg %>%
  ggplot() + 
  geom_histogram(aes(x = hwy, y = ..density..),
                 color = "black",
                 fill = "purple") +
  geom_density(aes(x = hwy),
               fill = "purple",
               alpha = 0.7) + 
  labs(x = "Highway MPG",
       y = "")

p2 <- mpg %>%
  mutate(
    drv = case_when(
      drv == "f" ~ "Forward Drive",
      drv == "r" ~ "Rear Drive",
      TRUE ~ "Four Wheel"
    )
  ) %>%
  ggplot() +
  geom_boxplot(aes(x = cty, y = drv, fill = drv),
               show.legend = FALSE) +
  labs(x = "City MPG",
       y = "")

p3 <- mpg %>%
  count(class) %>%
  ggplot() +
  geom_col(aes(x = fct_reorder(class, n), y = n)) +
  labs(x = "",
       y = "Count") + 
  coord_flip()

p4 <- mpg %>%
  ggplot() + 
  geom_point(aes(x = cty, y = hwy)) + 
  labs(x = "City MPG", 
       y = "Highway MPG")

pwk <- (p3 + p1) / (p2 + p4)
pwk + plot_annotation(
  title = "Some Plots About Cars"
)
```

## A Bad Viz Doesn't

```{r}
#| echo: false
#| message: false
#| warning: false

p1 <- mpg %>%
  ggplot() + 
  geom_bar(aes(x = displ)) + 
  labs(x = "Engine Displacement",
       y = "")

p2 <- mpg %>%
  ggplot() +
  geom_point(aes(x = as.factor(year), y = class)) +
  labs(x = "Year",
       y = "")

p3 <- mpg %>%
  ggplot() +
  geom_point(aes(x = cyl, y = cty)) +
  labs(x = "",
       y = "")

pwk <- p2 + (p1 / p3)
pwk + plot_annotation(
  title = "These Plots Stink!"
)
```

## Suggested Viz Choices

+ Single Numerical Variable

  + Histogram, Boxplot, or Density

+ Single Categorical Variable

  + Bar Graph

+ Two Numerical Variables

  + Scatterplot or Heatmap

+ Two Categorical Variables

  + Bar Graph with Fill Color or Heatmap

+ One Numerical and One Categorical Variable

  + Side-by-Side boxplots, overlayed or faceted histograms/densities
  
## Structure of a `ggplot()`

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1|2|3|4|1-4"

data %>% #the data to be plotted
  ggplot() + #initialize a plot
  geom_PLOT_TYPE(aes(...)) + #plot type and required mappings
  labs() #labels for plot
```

+ Note we can pipe (`%>%`) a data frame into a plot

  + We don't need a data frame though

+ Once we use `ggplot()` we use `+` to add layers instead of piping
+ `geom_*()` layers require *aesthetics* to map variables to plot features

  + Different *geoms* have different required/permitted aesthetics

+ Can add multiple *geoms* to a single plot
+ Every plot should include labels
+ Find [a `ggplot` cheatsheet here](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf)

## Let's Do This!

1. Open RStudio
2. Check the top-right corner, next to the translucent blue box icon to verify that you are working in your `MAT434` project space that is managing your GitHub Repo.

    + If you see `None` there instead of your project name, open your project by navigating to `File -> Open Project` or by using the dropdown menu near the project box
    
3. Open your Quarto Notebook on the MLB batted balls and home runs from last time

4. Add a section on data visualizations

## Suggested Viz Choices, Revisited

+ Single Numerical Variable

  + `geom_boxplot()`, `geom_histogram()`, or `geom_density()`
  
    + Require `x` *or* `y` aesthetic (but not both!)
    + For example, `geom_density(aes(x = hwy))`
  
+ Single Categorical Variable

  + `geom_bar()`
  
    + Requires `x` *or* `y` aesthetic (but not both!)
    + For example, `geom_bar(aes(x = class))`
  
  + `geom_col()`
    
    + Can have both `x` and `y` aesthetic
    + Example, `geom_col(aes(x = class, y = n))`
    
## Suggested Viz Choices, Revisited

+ Two Numerical Variables

  + `geom_point()` or `geom_hexbin()`
  
    + Require both `x` *and* `y` aesthetic
    + For example, `geom_point(aes(x = cty, y = hwy))`
  
+ Two Categorical Variables

  + `geom_bar()`
  
    + Use `x` *and* `fill` aesthetics
    + For example, `geom_bar(aes(x = class, fill = drv))`

## Suggested Viz Choices, Revisited

+ One Numerical and One Categorical Variable

  + `geom_boxplot()`
  
    + Use both `x` and `y` aesthetics
    + For example, `geom_boxplot(aes(x = hwy, y = class))`
  
  + `geom_density()` or `geom_histogram()`
 
    + Use only `x` aesthetic
    + Add layer `facet_wrap(~ VAR_NAME)`

+ Other available *aesthetics* include `color`, `size`, `shape`, and `alpha` (transparency)

  + Remember, specific *geoms* permit only specific *aesthetics*
  
## Try It!

+ Start with a few single-variable plots

  + Plot the *distribution* of home-runs versus non-home-runs (`is_home_run`)
  + Plot the *distribution* of `pitch_mph`
  + Plot the *distribution* of `launch_speed`
  + Plot the *distribution* of `launch_angle`

+ Ask ChatGPT (or your favorite LLM) to help you troubleshoot your plots or to "trick out" at least one

## Try It!

+ Now build some multivariable plots

  + you might focus on searching for features in your data set that seem associated with whether or not a batted ball was a home run.
  + if you are interested in potential other associations, investigate those as well
  
+ Again, ask your favorite LLM for help troubleshooting code or "tricking out" your plots.

## On Labels

+ The `labs()` layer permits global plot labels and labels for any mapped aesthetic

  + `title`
  + `subtitle`
  + `caption`
  + `alt` (for alt-text)
  + `x`
  + `y`
  + `color`
  + `fill`
  + etc.
  
## Organizing Plots with `{patchwork}`

+ Often you'll want to arrange plots together, rather than printing them out one at a time
+ The `{patchwork}` package provides very easy and intuitive framework for doing this.

1. Create each of your plots, but store them into variables `p1`, `p2`, ...
2. Use `+` to organize plots side-by-side, and `/` to organize plots over/under one another.

    + For example, `(p1 + p2) / p3` will arrange plots `p1` and `p2` side-by-side, with plot `p3` underneath them.
    
## Try It!

+ Try storing some of your earlier plots and printing them out in a combined plot by using `{patchwork}`
+ Keep playing around -- post your best plot to the `#how-do-i` channel on Slack

