---
title: '`tidymodels` Framework Example Model Construction`'
output:
  html_document:
    theme: cerulean
    highlight: pygments
  pdf_document: default
---

```{r global-options, include=FALSE}
#knitr::opts_chunk$set(eval = FALSE)
library(tidyverse)
library(tidymodels)
library(kableExtra)
```

**Objectives:** This notebook shows an example of a model being assessed and constructed using the framework proposed in notebook `x_ModelingFrameworkAndTidymodelsReview`. We'll use the `penguins` data from the `palmerpenguins` packages and try to predict penguin `species` using the other variables available in the dataset.

## The `penguins` data

The `penguins` data set contains measurements on `r penguins %>% nrow()` penguins from the Palmer Research Station in Antarctica. Measurements are mostly physiological, with some demographic information including `sex`, `year` observed, and `island` of observation.

```{r}
penguins <- palmerpenguins::penguins %>%
  filter(!is.na(species))

penguins %>%
  head(n = 2) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

In this notebook, we'll show how to assess, fit and use a model to classify observed penguin `species` given physiological attributes. To keep things simple, we'll use `body_mass_g`, `island`, `sex`, and `flipper_length_mm` as predictors.

## Splitting Data and Creating Folds

We'll start by splitting our data into *training* and *test* sets and create folds using our *training* data.

```{r}
set.seed(123)
penguin_splits <- initial_split(penguins, prop = 0.75)
train <- training(penguin_splits)
test <- testing(penguin_splits)

set.seed(456)
train_folds <- vfold_cv(train, v = 3)
```

## Building Model Workflows

We'll build workflows for a logistic regression model and a decision tree model.

```{r}
#knn workflow
knn_spec <- nearest_neighbor() %>%
  set_engine("kknn") %>%
  set_mode("classification")

knn_rec <- recipe(species ~ body_mass_g + island + sex + flipper_length_mm, data = train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

knn_wf <- workflow() %>%
  add_model(knn_spec) %>%
  add_recipe(knn_rec)

#decision tree workflow
dt_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

dt_rec <- recipe(species ~ body_mass_g + island + sex + flipper_length_mm, data = train) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors())

dt_wf <- workflow() %>%
  add_model(dt_spec) %>%
  add_recipe(dt_rec)
```

## Evaluating Models Using Cross-Validation

We'll now use cross-validation to evaluate our two workflows.

```{r}
knn_cv_results <- knn_wf %>%
  fit_resamples(train_folds)

dt_cv_results <- dt_wf %>%
  fit_resamples(train_folds)

knn_cv_results %>%
  collect_metrics() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))

dt_cv_results %>%
  collect_metrics() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

It looks like the nearest neighbor model outperforms the decision tree using both *accuracy* and *ROC-AUC* as the performance metric. Let's fit the nearest neighbors model to the training data.

## Fit, Assess, and Utilize Our Best Model

We'll start by fitting our best model to our training data.

```{r}
knn_fit <- knn_wf %>%
  fit(train)
```

Now that the model has been fitted, we'll use our model to make predictions on our `test` observations. The performance metrics should align with our cross-validation estimate (if not, then we'll have concerns which will prevent us from putting the model into production).

```{r}
knn_fit %>%
  augment(test) %>%
  accuracy(species, .pred_class)
```

The model was even more accurate than expected given the cross-validation results. Since it doesn't look like our model is overfit, let's use it to make a prediction on the `species` of a newly observed penguin.

```{r}
new_penguin <- tibble(
  body_mass_g = 4750,
  island = "Dream",
  sex = "male",
  flipper_length_mm = 38
)

new_penguin %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))

knn_fit %>%
  augment(new_penguin) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("hover", "striped"))
```

## Summary

There it is -- we've walked through a basic modeling process using the `tidymodels` framework. We'll discuss this in greater detail and introduce more advanced techniques throughout MAT434.
